---
title: "Exploring variation in scRNA-seq datasets"
output: html_notebook
---



# General
Load necessary libraries
 
```{r}
# Package installation
## CRAN
###install.packages(c("Seurat", "Matrix", "dplyr", "gprofiler2", "ggplot2", 
###                   "patchwork", "gplots", "cowplot", "pheatmap", 
###                     "reshape2", "devtools"))
## Github
### devtools::install_github('satijalab/seurat-data')

# data manipulation
library(Seurat)
library(SeuratData)
library(Matrix)
library(dplyr)
library(reshape2)

# plotting
library(ggplot2)
library(patchwork)
library(cowplot)
library(pheatmap)

set.seed(1)
```



# Exploring variation
## Variation by selected variable genes
Load data

```{r}
srat = readRDS("Example_filt_srat.RDS")
```

Normalise expression of each gene in each cell by the total counts in that cell, with a scaling factor of 10.000, and then log-transform the data.

```{r}
DefaultAssay(srat) = "RNA"
srat = NormalizeData(srat, normalization.method = "LogNormalize", scale.factor = 10000)
```

Run with different numbers of variable genes

```{r}
nfeature_l = c(100, 1000, 5000, 12000)
srat_l = list()
for(i in nfeature_l){
  start.time = Sys.time()
  print(paste0("Using the top ", i, " variable genes."))
  
  # calculate variable features (with different top genes)
  srat_l[[paste0("f", i)]] = FindVariableFeatures(srat, 
                                                  selection.method = "vst", 
                                                  nfeatures = i, verbose = F)
  # scale the data
  srat_l[[paste0("f", i)]] = ScaleData(srat_l[[paste0("f", i)]], 
                                       vars.to.regress = c("nCount_RNA"), 
                                       do.scale = F, verbose = F)
         
  # run PCA
  srat_l[[paste0("f", i)]] = RunPCA(srat_l[[paste0("f", i)]], verbose = F, 
                                    assay = "RNA", npcs = 30) # asking for fewer to speed up execution
         
  # run UMAP
  ncomp = 15
  srat_l[[paste0("f", i)]] = RunUMAP(srat_l[[paste0("f", i)]], 
                                           dims = 1:ncomp, verbose = F)
  
  # clustering
  red = "pca"
  srat_l[[paste0("f", i)]] = FindNeighbors(srat_l[[paste0("f", i)]], dims = 1:ncomp, 
                                                 force.recalc = T, verbose = F,
                                                 reduction = red, graph.name = paste0(red, ncomp))
  srat_l[[paste0("f", i)]]  = FindClusters(srat_l[[paste0("f", i)]] , algorithm = 2, 
                                                 verbose = F, graph.name = paste0(red, ncomp),
                                                 resolution = 3) # using a hig resolution for cluster comparisons
  srat_l[[paste0("f", i)]] = SetIdent(srat_l[[paste0("f", i)]], value = paste0(red, ncomp, "_res.3"))
  end.time = Sys.time()
  print(paste0("Execution time: ", round(difftime(end.time, start.time, units = "sec"), 1), " seconds."))
  gc(full = T)
}
```



```{r, fig.height=8, fig.width=8}
plt_list = list()
# iterate each clustering resolution
for(g in names(srat_l)){
  plt_list[[g]] = DimPlot(srat_l[[g]], reduction = "umap", label = T, 
                          raster = F, pt.size = 0.2, shuffle = T)+
    labs(subtitle = g)+
    theme(legend.position = "none",
          aspect.ratio = 1)
}
cowplot::plot_grid(plotlist = plt_list, ncol = 2)
```

Compare clusters by correlation

```{r}
# choose a common set of genes to calculate the correlations (marker genes would likely work best)
g = VariableFeatures(srat_l$f5000)
# all pairwise combinations
comp_obj = combn(names(srat_l), 2)
for(i in 1:ncol(comp_obj)){
  # calculate avg expression per cluster for each set of genes
  avg1 = AverageExpression(srat_l[[comp_obj[1,i]]], assay = "RNA", features = g, group.by = "pca15_res.3")$RNA
  avg2 = AverageExpression(srat_l[[comp_obj[2,i]]], assay = "RNA", features = g, group.by = "pca15_res.3")$RNA
  
  # correlation (Spearman)
  cor_mat = cor(as.matrix(avg1), as.matrix(avg2), method = "sp")
  
  pheatmap::pheatmap(cor_mat, clustering_method = "ward.D", 
                     main = paste0(comp_obj[1,i], " (rows) vs ", comp_obj[2,i], " (columns)"))
}
```


## Variation in principal components
Load data

```{r}
srat = readRDS("Example_final_sct.RDS")
```

Check which PCs influence which clusters

```{r, fig.height=8, fig.width=15}
pca_dat = data.frame(Reductions(srat, "pca")@cell.embeddings[,1:18])
pca_dat$cl = srat$pca15_res.1
plot_df = reshape2::melt(pca_dat)

ggplot(plot_df, aes(x = cl, y = value, colour = cl))+
  facet_wrap(~variable, scales = "free", ncol = 6)+
  ggbeeswarm::geom_quasirandom(size = 0.2)+
  geom_hline(yintercept = 0)+
  theme_classic()+
  theme(legend.position = "none")
```

Now we will remove the PCs tat seem more associated with cluster 8

```{r}
pcs_use = c(1:11, 13)
srat_rep = Seurat::RunUMAP(srat, reduction = "pca", verbose = F, 
                           dims = pcs_use, reduction.name = "umap_rep")
srat_rep = FindNeighbors(srat_rep, dims = pcs_use, force.recalc = T, 
                         verbose = F, reduction = "pca", graph.name = "pcarec")
srat_rep = FindClusters(srat_rep, algorithm = 2, verbose = F, graph.name = "pcarec",
                        resolution = 1)
```

And plot the original clustering, with the old and new UMAPs

```{r}
plt_before = DimPlot(srat_rep, reduction = "umap", 
                     group.by = "pca15_res.1", label = T)
plt_after = DimPlot(srat_rep, reduction = "umap_rep", 
                    group.by = "pca15_res.1", label = T)

plt_before+plt_after
```









